---
title: African Mammals
author: Urs Kalbitzer
date: '2020-01-11'
tags:
  - IUCN
  - R
  - Red List
  - rnaturalearth
  - rpatial
  - sf
slug: african-mammals-map
lastmod: '2020-01-11T16:55:58-05:00'
featured: no
image:
  placement: 1
  caption: ''
  focal_point: Center
  preview_only: no
---

Interested in the distribution of animal species on a global or continental level? This can be easily illustrated with R using the data provided by the [IUCN Red List](https://www.iucnredlist.org). In this post, I will focus on mammals in Africa, the continent where I have conducted most of my field work. But the code below can be easily adapted to other areas of the world and other taxonomic groups of species, or only species at low or high risk of extinction (though the completeness of the IUCN data is variable across different groups).    

In this post, I describe the following steps:

1. Download and process the _IUCN Red List_ spatial data set for terrestrial mammals
2. Download a map of Africa
3. Create a hexagon grid for Africa
4. Combine species ranges with the grid and plot the map

The processing and summarizing of the spatial data is mostly done with the impressive `sf` package. For my first attempts to work with spatial data in R (more specifically with vector data), I used the `sp` package. But the `sf` package, which is the successor of the `sp`, is much easier to use. Additionally, it integrates some of the [`tidyverse`](https://www.tidyverse.org) functions, most importantly many of the `dplyr` functions, which so many R users are already using for other purposes. Also, the `sf` package is very well documented and the vignettes are extremely helpful. They can be found on the [github page of the package](https://github.com/r-spatial/sf) or by typing `vignette(package = "sf")`. I also would like to mention the book ['Geocomputation with R'](https://geocompr.robinlovelace.net), which is a very informative (open source) resource for anybody interested in working with spatial data in R.

## 1. Download and process the _IUCN Red List_ spatial data set for terrestrial mammals

### 1.1 Prepare R
```{r prepare_r, message = FALSE}
rm(list = ls())
library(tidyverse)
library(sf)
```

### 1.2 Download and process the _IUCN_ terrestial mammals shapefile

The _IUCN Red List_ spatial data can be obtained [here](https://www.iucnredlist.org/resources/spatial-data-download). The _IUCN_ datasets are freely available for non-commercial use, but they have to be downloaded manually because you have to register for an account and provide a description of the intended usage. For academic use, you can (usually) download the dataset immediately after you made your request.  

The ranges of populations and species are all provided as polygons in the downloadable shapefiles. However, some species might be missing, and some ranges might be incomplete. Nevertheless, it is an impressive data set, and at least for primates, it looks fairly complete. Further details (and limitations) can be found on the _IUCN_ spatial data webpage.  

For this post, I downloaded the "Terrestrial Mammal" datasets, which is about ~ 600 MB. Then, I moved the downloaded folder ('TERRESTRIAL_MAMMALS') into the folder "~/Sync/iucn_data/" on my harddrive and renamed it to 'TERRESTRIAL_MAMMALS_2020_01_11'.  

I split up the following process into two steps: 1) load, simplify (_i.e._, reduce the size of the object), and save the processed shapefile a .gpkg file; and 2) re-load the processed files. The first step will not be evaluated in the final notebook so that the 600 MB shapefile does not have to be processed every time I run this notebook.
```{r process_raw_shapefile, eval = FALSE}
mammals <- st_read("~/Sync/iucn_data/TERRESTRIAL_MAMMALS_2020_01_11")

# Simplify the polygons to reduce the size of the object/file. This may take a
# few moments but the size of the object (in the R environment) changes from 906
# Mb to 8.4 Mb (with dTolerance = 5), which is much easier to handle.
mammals_simple <- mammals %>% 
  st_simplify(dTolerance = 5)

mammals_simple <- mammals_simple %>% 
  select(binomial, presence,legend, category)

# Write the object to a file and clean up. I use .gpkg extension here because it
# just requires one file instead of several files created for shapefiles. Also,
# it's an open format and it appears to be just as good as shapefiles (at least
# for everything I've done so far). Look here for more information:
# https://www.gis-blog.com/geopackage-vs-shapefile/
st_write(mammals_simple, "data/iucn_mammal_distribution.gpkg")
rm(list = c("mammals", "mammals_simple"))
```

_Note:_ I set the `dTolerance` argument to 5, but for a more precise map, it might be good to change it to 0.1 or a lower value.  
Also, I only kept four of the columns: `binomial` (the species name), `presence` and `legend` (indicating whether a population is still present), and `category` (the risk category for each species). For other projects, additional columns can be added. For example, `order_` or `family` indicate the taxonomic order/family of a species.    

Now, load the file (step 2)
```{r}
mammals <- st_read("data/iucn_mammal_distribution.gpkg")
```



Quick look at the sf object:
```{r}
head(mammals)
```

For an explanation of sf objects, look at the [first vignette of the sf package](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html). Every row in this data frame represents one population of species `binomial`, with additional information about the presence and risk category (columns `presence`, `legend`, and `category`). The sf-column `geom` contains the geographic information describing the range of each population. As far as I understand this data set, several populations might be combined into one row (the sf-column can contain several polygons), but in other cases, each population has its own row.  

We can simply plot all polygons (for all rows) with:
```{r}
plot(st_geometry(mammals))
```

This plot shows the distributions of all mammal species/populations, but it also includes populations that are no longer present. Let's have a look at the different codes for `presence` and the description, which can be found in the column `legend`.

```{r}
mammals %>%
  st_set_geometry(NULL) %>%
  distinct(presence, legend) %>% 
  arrange(presence)
  # knitr::kable()
```

Codes/numbers 1, 2, and 3 refer to 'extant', 'probably extant', and 'possibly extant' populations, and I only keep these populations here.
```{r}
mammals_extant <- mammals %>% 
  filter(presence %in% c(1,2,3))
```

## 2. Download a map of Africa

In R, it's very easy to get maps of countries or continents from https://www.naturalearthdata.com using the `rnaturalearth` package. With the `ne_countries()` function, all countries (as polygons) for the specified continent can be downloaded. Here, I am only interested in a map of the whole continent, not single countries. This can be achieved with the `dplyr::summarize()` function, which 'summarizes' the polygons in all rows in the sf object (the countries) to one polygon (the continent). One of the examples showing how nicely the tidyverse functionality is integrated into the `sf` package!

```{r}
africa_map <- rnaturalearth::ne_countries(continent = "Africa",
                                          returnclass = "sf") %>%
  st_set_precision(1e9) %>%
  summarize
{
par(mar = c(0, 0, 0, 0))
plot(st_geometry(africa_map))
}
```


## 3. Create a hexagon grid for Africa

Since I want to illustrate how species richness varies across regions, the number of species have to be summarized for some kind of (spatial) subsets of Africa. For example, this could be done by country. But here, I use a grid of equally sized hexagon cells to illustrate the varying species richness across the continent. This is possible with the `sf::st_make_grid()` function. The argument `square = F` specifies that hexagons are created instead of squares. Furthermore, I create a `grid_id` column, which will be required as identifier for grid cells further below.

```{r, message = FALSE, warning = FALSE}
africa_grid <- st_make_grid(africa_map,
                            what = "polygons",
                            cellsize = 0.75,
                            square = F) %>% 
  st_sf() %>% 
  mutate(grid_id = row_number())
```

_Note_: the cell size can be adjusted by changing the `cellsize` argument, but this will also affect the number of species per cell. Larger cells are more likely to intersect with more species than smaller cells.

For the map, I only want to keep the parts of the grid cells that are on the continent. This might be problematic in some cases because the cells at the edge of the continent are smaller than cells within the continent. But since I only summarize terrestrial mammals here, I think it might also be misleading if the cells cover non-terrestrial area without including marine mammals.

```{r, message = FALSE, warning = FALSE}
africa_grid_clipped <- st_intersection(africa_grid, africa_map)

{
plot(africa_map$geometry, reset = F, axes = T)
plot(st_geometry(africa_grid_clipped), color = "white",
     add = T, border = rgb(0, 0, 0, 0.3))
}
```

## 4. Combine species ranges with the grid and plot the map

### 4.1 Only keep population ranges within Africa

With `st_intersection()`, only the portion of population ranges within Africa are kept.
Then, ranges are 'summarized' by species, which means that for species with several rows in the data frame, these rows are combined into one row with a (multi) polygon describing the entire range of this species. The use of `dplyr::group_by()` in combination with `dplyr::summarize()` with an sf object is another great example for the integration of the tidyverse functionality in the sf package.

```{r, message = FALSE, warning = FALSE}
africa_mammals <- st_intersection(mammals, africa_map) %>% 
  group_by(binomial) %>% 
  summarize()
{
par(mar = c(2,2,2,2))
plot(st_geometry(africa_mammals), axes = T)
}
```

### 4.2 Combine ranges with the grid

The range polygons can be combined with the grid map using the `sf::st_join()` function. Then, the number of species is counted per grid cell using the `grid_id` column created above.
```{r, message = FALSE, warning = FALSE}
# This may take a few minutes
species_per_cell <- africa_grid_clipped %>% 
  st_join(africa_mammals)

species_per_cell_sums <- species_per_cell %>% 
  group_by(grid_id) %>% 
  summarize(species_n = n())
```

### 4.3 Create the plot
The standard output with `plot()` looks like this.
```{r}
plot(species_per_cell_sums["species_n"])
```

The plot created with `plot()` can be customized, but I find it easier to customize plots with the `ggplot2` package. Therefore, I create another plot with `ggplot()` using a different color palette, and applying some other modifications.

```{r}
african_mammals_map <- ggplot() +
  geom_sf(data = species_per_cell_sums,
          aes(fill = species_n),
          size = 0) +
  scale_fill_gradient2(name = "Number of\nSpecies",
                       low = "#004529",
                       mid = "#f7fcb9",
                       high = "#7f0000",
                       midpoint = max(species_per_cell_sums$species_n)/2) +
  geom_sf(data = africa_map, fill = NA) +
  labs(title = "Mammal Species in Africa") +
  theme_void() +
  theme(legend.position = c(0.1, 0.1),
        legend.justification = c(0, 0),
        plot.title = element_text(hjust = .5))
african_mammals_map
```

```{r, include = FALSE}
plot_featured <- african_mammals_map +
  theme(legend.position = "none",
        plot.title = element_blank(),
        plot.margin = margin(c(0, 0, 0, 0)))
cowplot::save_plot("featured.jpg", plot_featured)
```
